Cognate beginnings to bilingual lexical growth

### 2023-05-08 _appendix/_syllable-frequency.qmd
Fix phonological transcriptions and render manuscript

````{r}
duplicated_te <- bvq_data$pool$te[duplicated(bvq_data$pool$te)]

pool_tmp <- bvq_data$pool |>
    # drop items with missing observations in these variables
    drop_na(sampa, wordbank_lemma) |>
    mutate(xsampa = ipa(ipa)) |> 
    filter(n_lemmas == 1,
           # exclude items with more than two lemmas
           !is_multiword,
           # exclude multi-word items
           include,
           # exclude problematic items (e.g., multi-word items)
           te %in% duplicated_te) |>
    add_count(te, name = "n_te") |> # get only items with one translation in each language
    filter(n_te == 2) |>
    distinct(language, te, .keep_all = TRUE) |> 
    select(language, te, item, label, childes_lemma, xsampa) |> 
    mutate(syll = strsplit(xsampa, 
                           split = paste0(.sep = c("\\.", "\\\""), 
                                          collapse = "|")),
           syll = lapply(syll, function(x) x[x != ""]),
           n_syll = map_int(syll, length),
           item = str_remove(item, "cat_|spa_"))

@@

```{r items-table}
#| label: tbl-items
#| tbl-cap: Sample of items included in the BVQ questionnaire and their syllabified SAMPA transcriptions in Catalan and Spanish
items |> 
    distinct(te, item, language, meaning, xsampa, n_syll) |> 
    left_join(distinct(pool_tmp, te, language, item, label),
              by = join_by(te, item, language)) |> 
    pivot_wider(names_from = language,
                values_from = c(item, label, xsampa, n_syll),
                id_cols = c(te, meaning),
                names_repair = make_clean_names,
                values_fn = first) |> 
    sample_n(15) |> 
    gt() |> 
    cols_hide(c(te, item_catalan, item_spanish)) |> 
    tab_spanner("Catalan", matches("catalan")) |> 
    tab_spanner("Spanish", matches("spanish")) |> 
    cols_label(meaning = "Translation",
               label_catalan = "Item",
               xsampa_catalan = "X-SAMPA",
               n_syll_catalan = "Syllables",
               label_spanish = "Item",
               xsampa_spanish = "X-SAMPA",
               n_syll_spanish = "Syllables") |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels())
```

@@

```{r apafg-syl-number}
#| label: apafg-syl-number
#| apa-cap: Distribution of the number of syllables in Catalan and Spanish
#| fig-width: 7
#| fig-height: 3
pool_tmp |> 
    group_by(language) |> 
    add_count(language, name = "n_lang") |> 
    ungroup() |> 
    count(language, n_lang, n_syll) |> 
    mutate(prop = n/n_lang) |> 
    ggplot(aes(n_syll, n)) +
    facet_wrap(~ language) +
    geom_col(colour = "white", fill = "grey70") +
    geom_text(aes(label = percent(prop)),
              size = 3,
              position = position_nudge(y = 15),
              colour = "black") +
    labs(x = "Number of syllables",
         y = "Number of words",
         fill = "Language") +
    theme(legend.position = "none",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
```

@@

```{r apafg-syl-freq}
#| label: apafg-syl-freq
#| apa-cap: "Distribution of apositional syllable frequencies in Spanish and Catalan"
#| fig-height: 3.5
#| fig-width: 7
syll <- items |> 
    left_join(distinct(pool_tmp, te, childes_lemma),
              by = join_by(te)) |> 
    left_join(select(childes, token, freq_lex = freq_million),
              by = c("childes_lemma" = "token")) |> 
    select(language, item, te, meaning, xsampa, lv, freq_lex, syll) |> 
    unnest_longer(syll) |> 
    group_by(language, syll) |> 
    summarise(freq_syll = sum(freq_lex),
              .groups = "drop") |> 
    arrange(language, desc(freq_syll))

@@

```{r apafg-top-syll-freq}
#| label: apafg-top-syll-freq
#| apa-cap: "Top 20 most frequent syllables"
#| fig-width: 8
#| fig-height: 3
#| out-width: 80%
syll |> 
    group_by(language) |> 
    top_n(20, freq_syll) |> 
    ggplot(aes(reorder_within(syll, desc(freq_syll), language),
               freq_syll)) +
    facet_wrap(~ language,
               scales = "free_x",
               ncol = 1) +
    geom_col(colour = "white",
             fill = "grey70") +
    labs(x = "Syllable",
         y = "Relative frequency\n(counts per million)",
         colour = "Language",
         fill = "Language") +
    scale_x_reordered() +
    scale_y_continuous(labels = \(x) format(x, big.mark = ",")) +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 10),
          panel.grid = element_blank())
```

@@

```{r syllables-posterior}
rope_interval <- c(lower = -0.5, upper = 0.5)
# tidy predictor names
str_repl <- c(
    "b_Intercept" = "Intercept",
    "b_n_syll_std" = glue("Syllables (+1 SD, {round(sd(syllables_data$n_syll), 3)})"),
    "b_lv_std" = glue("Cognateness (+1 SD, {percent(sd(syllables_data$lv), accuracy = 0.01)})")
)

>>>>

```{r prepare-syllables}
#| label: prepare-syllables
#| echo: false
#| warning: false
#| message: false
duplicated_te <- bvq_data$pool$te[duplicated(bvq_data$pool$te)]

pool_tmp <- bvq_data$pool |>
    # drop items with missing observations in these variables
    drop_na(xsampa, wordbank_lemma) |>
    filter(n_lemmas == 1,
           # exclude items with more than two lemmas
           !is_multiword,
           # exclude multi-word items
           include,
           # exclude problematic items (e.g., multi-word items)
           te %in% duplicated_te) |>
    add_count(te, name = "n_te") |> # get only items with one translation in each language
    filter(n_te == 2) |>
    distinct(language, te, .keep_all = TRUE) |> 
    select(language, te, item, label, childes_lemma, xsampa) |> 
    mutate(syll = strsplit(xsampa, 
                           split = paste0(.sep = c("\\.", "\\\""), 
                                          collapse = "|")),
           syll = lapply(syll, function(x) x[x != ""]),
           n_syll = map_int(syll, length),
           item = str_remove(item, "cat_|spa_"))

@@

```{r items-table}
#| label: tbl-items
#| tbl-cap: Sample of items included in the BVQ questionnaire and their syllabified SAMPA transcriptions in Catalan and Spanish
#| echo: false
#| warning: false
#| message: false
items |> 
    distinct(te, item, language, meaning, xsampa, n_syll) |> 
    left_join(distinct(pool_tmp, te, language, item, label),
              by = join_by(te, item, language)) |> 
    pivot_wider(names_from = language,
                values_from = c(item, label, xsampa, n_syll),
                id_cols = c(te, meaning),
                names_repair = make_clean_names,
                values_fn = first) |> 
    sample_n(15) |> 
    gt() |> 
    cols_hide(c(te, item_catalan, item_spanish)) |> 
    tab_spanner("Catalan", matches("catalan")) |> 
    tab_spanner("Spanish", matches("spanish")) |> 
    cols_label(meaning = "Translation",
               label_catalan = "Item",
               xsampa_catalan = "X-SAMPA",
               n_syll_catalan = "Syllables",
               label_spanish = "Item",
               xsampa_spanish = "X-SAMPA",
               n_syll_spanish = "Syllables") |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels())
```

@@

```{r apafg-syl-number}
#| label: apafg-syl-number
#| apa-cap: Distribution of the number of syllables in Catalan and Spanish
#| fig-width: 7
#| fig-height: 3
#| echo: false
#| warning: false
#| message: false
pool_tmp |> 
    group_by(language) |> 
    add_count(language, name = "n_lang") |> 
    ungroup() |> 
    count(language, n_lang, n_syll) |> 
    mutate(prop = n/n_lang) |> 
    ggplot(aes(n_syll, n)) +
    facet_wrap(~ language) +
    geom_col(colour = "white", fill = "grey70") +
    geom_text(aes(label = percent(prop)),
              size = 3,
              position = position_nudge(y = 15),
              colour = "black") +
    labs(x = "Number of syllables",
         y = "Number of words",
         fill = "Language") +
    theme(legend.position = "none",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
```

@@

```{r apafg-syl-freq}
#| label: apafg-syl-freq
#| apa-cap: "Distribution of apositional syllable frequencies in Spanish and Catalan"
#| fig-height: 3.5
#| fig-width: 7
#| echo: false
#| warning: false
#| message: false
syll <- items |> 
    left_join(distinct(pool_tmp, te, childes_lemma),
              by = join_by(te)) |> 
    left_join(select(childes, token, freq_lex = freq_million),
              by = c("childes_lemma" = "token")) |> 
    select(language, item, te, meaning, xsampa, lv, freq_lex, syll) |> 
    unnest_longer(syll) |> 
    group_by(language, syll) |> 
    summarise(freq_syll = sum(freq_lex),
              .groups = "drop") |> 
    arrange(language, desc(freq_syll))

@@

```{r apafg-top-syll-freq}
#| label: apafg-top-syll-freq
#| apa-cap: "Top 20 most frequent syllables"
#| fig-width: 8
#| fig-height: 3
#| out-width: 80%
#| echo: false
#| warning: false
#| message: false
syll |> 
    group_by(language) |> 
    top_n(20, freq_syll) |> 
    ggplot(aes(reorder_within(syll, desc(freq_syll), language),
               freq_syll)) +
    facet_wrap(~ language,
               scales = "free_x",
               ncol = 1) +
    geom_col(colour = "white",
             fill = "grey70") +
    labs(x = "Syllable",
         y = "Relative frequency\n(counts per million)",
         colour = "Language",
         fill = "Language") +
    scale_x_reordered() +
    scale_y_continuous(labels = \(x) format(x, big.mark = ",")) +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 10),
          panel.grid = element_blank())
```

@@

```{r syllables-posterior}
#| label: syllables-posterior
#| echo: false
#| warning: false
#| message: false
rope_interval <- c(lower = -0.5, upper = 0.5)
# tidy predictor names
str_repl <- c(
    "b_Intercept" = "Intercept",
    "b_n_syll_std" = glue("Syllables (+1 SD, {round(sd(syllables_data$n_syll), 3)})"),
    "b_lv_std" = glue("Cognateness (+1 SD, {percent(sd(syllables_data$lv), accuracy = 0.01)})")
)
